Refactoring the `nls()` function in R

### 2022-06-02 RefactoringNLS/RefactoringNLS.Rmd
Add section on bounds

>>>>

## Feature: bounds on parameters

There are many situations where the context of a problem constrains the values of parameters.
For example, one of us was asked many years ago to estimate a model. One paramater returned
a negative value. The client complained "But that is supposed to be the number of grain elevators
in Saskatchewan". That is, the number should have been a positive integer, and likely less
than a few thousand.

`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
the manual states

> The algorithm = "port" code appears unfinished, and does not even check that the starting value is within the bounds. Use with caution, especially where bounds are supplied.

This is clearly unsatisfactory for general use. Inclusion of bounds on parameters in nonlinear
least squares computations is relatively straightforward, and it is part of the default method
for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
script, which we do not run here for reasons of space, shows that the solutions found are not
optimal for the Hobbs test problem, even in its scaled form.

```{r chkbound, eval=FALSE}
#### bounds with formula specification of problem
## Use the Hobbs Weed problem
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
weedframe <- data.frame(y=weed, tt=tt)
st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
## Unscaled model
wmodu <- y ~ b1/(1+b2*exp(-b3*tt))
## Scaled model
wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt))
## We can provide the residual and Jacobian as functions
# Scaled Hobbs problem
shobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
  # This variant uses looping
  if(length(x) != 3) stop("shobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  100.0*x[1]/(1+x[2]*10.*exp(-0.1*x[3]*tt)) - y
}

shobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-0.1*x[3]*tt)
  zz  <-  100.0/(1+10.*x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -0.1*x[1]*zz*zz*yy
  jj[tt,3]   <-   0.01*x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}

##----- Hobbsbounded unique code starts here ------
require(minpack.lm)
traceval<-FALSE
cat("Infeasible start test\n")
start1inf <- c(b1=4, b2=4, b3=4) # b3 OUT OF BOUNDS for next few tries
## Infeasible start! No warning message!
anlM2i <- try(nlsLM(wmodu, start=start1inf, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
                    trace=traceval))

# feasible start i.e. on or within bounds
start1<-st
anlM1 <- try(nlsLM(wmodu, start=start1, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
                    trace=traceval))
print(anlM1)

# Hobbs scaled problem with bounds, formula specification

require(nlsr)
anlxb1 <- nlxb(wmods, start=start1, data=weedframe, lower=c(0,0,0),
                  upper=c(2,6,3), trace=traceval)
print(anlxb1)

## nlsLM seems NOT to work properly with bounds
anlsLM1b <- nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
                 upper=c(2,6,3), trace=traceval)
print(anlsLM1b)
newst<-coef(anlsLM1b)
print(newst)
anlsLM1bx <- nlsLM(wmods, start=newst, data=weedframe, lower=c(0,0,0),
                   upper=c(2,6,3))
print(anlsLM1bx)

cat("Single number bounds:\n")
anlsLM1b1 <- try(nlsLM(wmods, start=start1, data=weedframe, lower=0,
                   upper=3))
print(anlsLM1b1)
anlsLM1b1a <- try(nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
                   upper=c(3,3,3)))
print(anlsLM1b1a)

anlxb1b1<- try(nlxb(wmods, start=start1, data=weedframe, lower=0,
                     upper=3))
print(anlxb1b1)

anlm1b <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
                  upper=c(2,6,3))
print(anlm1b)

# functional presentation of problem  -- seems to work 
anlm1bx <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
                   upper=c(2,6,3))
print(anlm1bx)

anlfb1bx <- nlfb(start=start1, resfn=shobbs.res, jacfn=shobbs.jac, lower=c(0,0,0),
                  upper=c(2,6,3))
print(anlfb1bx)
```

### Philosophical considerations

Bounds on parameters raise some interesting and difficult questions about how uncertainty
in parameter estimates should be computed or reported. That is, the traditional "standard
errors" are generally taken to imply symmetric intervals about the point estimate in which
the parameter may be expected to be found with some probability under certain assumptions.
Bounds change those assumptions drastically. Hence `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.
